### 2020-10-20 paper.Rmd
folder re-arrange

---
citation_package: natbib
bibliography:
- references.bib
- packages.bib
biblio-style: apalike
link-citations: yes
editor_options:
  chunk_output_type: console
output:
  bookdown::pdf_document2: default
output_dir: "docs"
---

@@

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
), "packages.bib")
```

@@

Short sprint speed is one of the most distinguishable and admired physical traits in sports, and one that have puzzled coaches and sports scientists from the time when two humanoids raced against each other (if they were available back then). Short sprints are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. This is approximately less than 6 seconds in duration (why? ATP/CP? --- REF), but the distance can vary based on the level of the athlete (i.e. lower level athletes can reach this point earlier, around 30-40m, while elite sprinters much later, around 50-60m --- REF).

Short sprints have been modeled using the mono-exponential equation \@ref(eq:velocity-time) originally proposed by @doi:10.1098/rspb.1927.0035, and more recently popularized by @greenePredictingSprintDynamics1986, @clarkNFLCombine40Yard2017, @chellyLegPowerHopping2001, @morinSimpleMethodComputing2019, @morinInterpretingPowerForceVelocityProfiles2016, @haugenPowerForceVelocityProfilingSprinting2020, @haugenSprintMechanicalVariables2019, @haugenSprintMechanicalProperties2020, @doi:10.1098/rspb.1927.0035, @haugenDifferenceStartImpact2012, and @samozinoSimpleMethodMeasuring2016:

@@

The parameters of the equation \@ref(eq:velocity-time) are \textit{maximum sprinting speed} (MSS; expressed in $ms^-1$) and \textit{relative acceleration} (TAU). Mathematically, TAU represents the ratio of MSS to initial acceleration (MAC; \textit{maximal acceleration}, expressed in $ms^-2$) \@ref(eq:maximal-acceleration).

@@

By derivating equation \@ref(eq:velocity-time), we can get equation for acceleration \@ref(eq:acceleration-time).

@@

By integrating equation \@ref(eq:velocity-time), we can get equation for distance \@ref(eq:distance-time).

@@

kinematics_long <- kinematics %>%
  pivot_longer(
    cols = c("distance", "velocity", "acceleration"),
    names_to = "variable"
  ) %>%
  mutate(
    variable = factor(
      variable,
      levels = c("distance", "velocity", "acceleration"),
      labels = c("Distance (m)", "Velocity (m/s)", "Acceleration (m/s/s)")
    )
  )

gg <- ggplot(
  kinematics_long,
  aes(x = time, y = value, color = Athlete)
) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~variable, scales = "free_y") +
  ylab(NULL) +
  xlab("Time (s)") +
  theme(legend.position = "top")

@@

If we plot acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get Acceleration-Velocity profile, which is, according to the mathematical model, linear.

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(10) +
  geom_line(alpha = 0.7) +
  ylab("Acceleration (m/s/s)") +
  xlab("Velocity (m/s)") +
  theme(legend.position = "top")

@@

# Estimation using `shorts` package

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at 0m, 10m, 20m, 30m, 40m), and (2) getting a velocity trace using radar gun (usually at around 100Hz). Estimation of the MSS and TAU parameters from the equation \@ref(eq:velocity-time) is performed in `shorts` package using non-linear least squares regression implemented in the `nls` function in the base R [@R-base] and `nlme` function in the `nlme` package [@R-nlme] for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$m/s^2$ performing 40m sprint with timing gates positioned at each 10m split. When it comes to split times, distance is a predictor, and time is outcome variable, thus the equation \@ref(eq:velocity-time) takes another form:

@@

$W$ in the equation \@ref(eq:velocity-time) represents Lambert's W function [@R-LambertW]. MSS and TAU parameters are estimated using `shorts::model_using_splits` function:

@@

# Generate times by using:
# round(predict_time_at_distance(distance = split_distance, MSS = 9, TAU = 1.3), 2)

@@

Maximal relative power (PMAX) from the output is estimated using $\frac{MSS \times MAC}{4}$, which disregards the air resistance (which we will deal with soon). `time_correction` and `distance_corection` parameters will be covered later in the article.

Besides providing *residual standard error* (RSE), `shorts` functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls` functions:

@@

Here is a plot of observed and model predicted split times, with the help of the `ggplot2` package [@R-ggplot2]:

@@

ggplot(df, aes(x = distance)) +
  theme_bw(10) +
  geom_line(aes(y = time)) +
  geom_point(aes(y = pred_time), color = "red") +
  xlab("Distance (m)") +
  ylab("Time (sec)")
```

@@

To estimate force production at distance or time (using `shorts::predict_force_at_distance` and `shorts::predict_force_at_time` functions), and later power production (using `shorts::predict_power_at_distance` and `shorts::predict_power_at_time` functions), one needs to take into account the air resistance. Air resistance (in Newtons) is estimated using `shorts::get_air_resistance` function, which takes velocity, body mass (in kg), body height (in meters), barometric pressure (in Torrs), air temperature (in Celzius), and wind velocity (in $ms^-1$) as parameters (please refer to @arsacModelingEnergetics100m2002, @samozinoSimpleMethodMeasuring2016, and @vaningenschenauCanCyclePower1991 for more information):

@@

When estimating force and power, one can set the air resistance parameters using `...`:

@@

The easiest way to get all kinematics for 0-6sec short sprints is to use `shorts::predict_kinematics` functions:

@@

Plotting the model predictions can be done once we convert data from wide to long with the help of `dplyr`[@R-dplyr], `tidyr` [@R-tidyr], and `tidyverse` [@R-tidyverse] packages:

@@

Sports scientists and coaches might be interested in finding distances and times where 90% of maximum sprinting speed is reached, or where peak power is within 90% range. To help finding these threshold, `shorts` package comes with `shorts::find_` family of functions:

@@

Short sprints are often times performed with a group of athlete (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles (i.e., for each individual), or utilize mixed-effects models. To perform mixed-effects models in `shorts` for split times, one can use `shorts::mixed_model_using_splits` function. To demonstrate this functionality, we will load the `split_times` dataset provided in the `shorts` package:

@@

To following plot contains kinematics for all athletes in `split_times` dataset. Please note that power calculation takes default parameters for each individual:

@@

ggplot(
  filter(df, distance < 40),
  aes(x = distance, y = value, group = athlete, color = athlete)
) +
  theme_bw(10) +
  facet_wrap(~name, scales = "free_y") +
  geom_line(alpha = 0.7) +
  ylab(NULL) +
  xlab("Distance (m)") +
  theme(legend.position = "top")
```

@@

Estimation of the short sprint profile using radar gun data takes time as predictor and velocity as an target or outcome variable. Thus the equation \@ref(eq:velocity-time) is used to estimate MSS and TAU.

Let's consider the same example of an athlete with MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$m/s^2$ performing 40m sprint with velocity estimated using radar run (in this case with 1Hz sampling rate).

@@

Mixed-effects model using radar data is done using `shorts::mixed_model_using_radar` function. To perform mixed model, let's load data that comes with `shorts` package.

@@

With the previous examples, estimation can look simple. Unfortunately, in real life things are a bit more *messy*, and this messiness can bias the estimated MSS and TAU parameters. In the next section, we will demonstrate how real life issues can introduce biases, but we will also present potential solutions within the `shorts` package.

@@

One source of messiness in the estimation using radar gun, beside measurement error, is the time synchronization, where zero velocity happens at $t=0$. Let's use our athlete and add and deduct 0.5second to simulate bad synchronization and its effect on estimated MSS and TAU.

@@

ggplot(plot_df, aes(x = value, y = velocity, color = `Sync issue`)) +
  theme_bw(10) +
  geom_line(alpha = 0.7) +
  xlab("Time (s)") +
  ylab("Velocity (m/s)")
```

@@

As can be seen from the example, it's the TAU estimate that is mostly affected by a bad synchronization of time with velocity. Solution implemented in `shorts` package involves estimation of the *time correction* parameters using the following equation:

@@

This model is utilizes using the `shorts::model_using_radar_with_time_correction` function:

@@

When it comes to mixed-model approach, time correction can be modeled as fixed effect or random effect using the `shorts::mixed_model_using_radar_with_time_correction` function.

@@

Let's imagine we have two twin brothers with same short sprint characteristics: MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$m/s^2$. Let's call them John and Jack. The are both performing 40m sprint using timing gates set at 5, 10, 20, 30, and 40m. The initial timing gate at the start serves the purpose of activating the timing system (i.e., when they cross the beam).

John is our *theoretical model*, while Jack is *street smart guy*, most likely playing soccer, and decides to move slightly back (i.e. 0.5m) and use body rocking to initiate the start. Let's see how their sprints differ.

@@

As expected, Jack performs better than John, even if they have same short sprint capabilities. Let's see how this affects the MSS and TAU estimates:

@@

As can be seen from the results, *cheating* (or not having consistent start) at the begin of the sprint yields biased estimates, particularly for the TAU, MAC and PMAX. Let's create a small simulation to estimate effects of cheating distance over combinations of MSS and MAC on their estimates as well as residuals.

Simulated splits are 5, 10, 20, 30, 40, and 50m, with MSS and MAC varying from 6 to 9, and cheating distance varying from 0 to 1m.

```{r}
sim_df <- expand.grid(
  MSS = c(6, 7, 8, 9),
  MAC = c(6, 7, 8, 9),
  cheat_distance = c(
    seq(0, 0.001, length.out = 20),
    seq(0.001, 0.01, length.out = 20),
    seq(0.01, 0.1, length.out = 20),
    seq(0.1, 1, length.out = 20)
  ),
  distance = c(5, 10, 20, 30, 40, 50)
)

sim_df <- sim_df %>%
  mutate(
    TAU = MSS / MAC,
    PMAX = MSS * MAC / 4,
    true_distance = distance + cheat_distance,
    true_time = shorts::predict_time_at_distance(true_distance, MSS, TAU),
    stolen_time = shorts::predict_time_at_distance(cheat_distance, MSS, TAU),
    time = true_time - stolen_time
  )

@@

Now when we have a simulation dataset, we can check the model estimates and predictions, given the cheating distance:

@@

# estimated parameters and predited time
model_df <- sim_df %>%
  group_by(MSS, TAU, cheat_distance) %>%
  do(pred_wrapper(.)) %>%
  ungroup()

@@

The following image demonstrates the effect of cheating distance on estimated MSS:

```{r}
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, cheat_distance) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(df, aes(x = cheat_distance, y = est_MSS, color = MAC_string)) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~MSS_string, scales = "free_y") +
  xlab("Cheat distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(legend.title = element_blank())
```

As can be seen from the image, MSS is underestimated as cheating distance increases. The following image demonstrates the effect of cheating distance on estimated MAC:

```{r}
# MAC
ggplot(df, aes(x = cheat_distance, y = est_MAC, color = MSS_string)) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Cheat distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(legend.title = element_blank())
```

MAC (and also TAU) are highly affected by the cheating distance, and from the figure we can notice that MAC is overestimated as cheat distance increases.

And finally, the following image demonstrates the effect of cheating distance on estimated PMAX:

```{r}
# PMAX
ggplot(df, aes(x = cheat_distance, y = est_PMAX, color = MSS_string)) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Cheat distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(legend.title = element_blank())
```

@@

Model residuals are also affected by cheating distance. The shape of residuals depends on splits utilized, but here we can see the effect of the cheating distance on the model residuals per split distance:

```{r}
# Residuals
model_df <- model_df %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2)),
    group = paste(MSS, MAC, cheat_distance)
  )

ggplot(model_df, aes(y = residuals, x = distance, color = cheat_distance, group = group)) +
  theme_bw() +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (sec)")
```

If we merge individual facets (i.e., combinations of MSS and MAC), we can get simpler images conveying issues with residuals when there is cheating at the start:

```{r}
ggplot(model_df, aes(y = residuals, x = distance, color = cheat_distance, group = group)) +
  theme_bw() +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (sec)")
```

Clearly, cheating and sprint start timing issues and inconsistencies can result in biased parameters and predictions. This is particularly important in testing and comparing test results implementing different starting techniques and timing initiations (**REFERENCE**). Since speed trait is one of the hardest to improve, the effects of start inconsistencies can mask effects of the training intervention. It is thus of crucial importance to take these into account, either by standardizing the start (which is always a must), or by implementing the following techniques.

### How to deal with the cheating or sprint start timing issues?

Besides standardizing sprint starts, particularly for longitudinal monitoring of the training effects, one can implement additional strategies to make model estimates more robust.

The simplest strategy is to used validity study results (i.e. comparing timing technique of interest to a gold standard measurement) and to use correction factors (**REFERENCE**). One way to implement correction is to use *time correction* in a same manner used with the radar gun synchronization issues. Thus, equation \@ref(time-distance) gets another form:

@@

If using validity studies, this time correction can be provided (i.e., 0.3-0.5seconds; *REFERENCE*), which is oftentimes the case, or it can me estimated as MSS and TAU is estimated. To estimate time correction, we use `shorts::model_using_splits_with_time_correction` function. Here is how we can estimate Jack parameters using these two methods:

```{r}
jack_profile_fixed_time <- shorts::model_using_splits(
  distance = split_times$distance,
  time = split_times$jack_time,
  time_correction = 0.3
)

@@

jack_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters),
  unlist(jack_profile_fixed_time$parameters),
  unlist(jack_profile_time_estimated$parameters)
)

rownames(jack_parameters) <- c(
  "John",
  "Jack - No corrections",
  "Jack - Fixed time correction",
  "Jack - Estimated time correction"
)

@@

In Jack's case, both fixed time correction and time correction estimation yield parameters closer to John's (i.e. true parameters).

Another model definition, which is novel approach implemented in the `shorts` packages, is to ulitize *distance correction*, besides time correction. Thus, equation \@ref(time-distance) gets another form:

@@

jack_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters),
  unlist(jack_profile_fixed_time$parameters),
  unlist(jack_profile_time_estimated$parameters),
  unlist(jack_profile_distance_correction$parameters)
)

>>>>

---
documentclass: jss
author:
  - name: Mladen Jovanović
    affiliation: |
      | University of Belgrade
    address: |
      | Faculty of Sport and Physical Education
      | Belgrade, Serbia
    email: \email{coach.mladen.jovanovic@gmail.com}
  - name: Jason D. Vescovi
    affiliation: |
      | University of Toronto
    address: |
      | Faculty of Kinesiology and Physical Education
      | Graduate School of Exercise Science
      | Toronto, ON Canada
    email: \email{vescovij@gmail.com}
title:
  formatted: "\\pkg{shorts}: An R Package for Modeling Short Sprints"
abstract: >
  The abstract of the article.
keywords:
  formatted: [sprints, sports, modeling]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
citation_package: natbib
bibliography: [references.bib, packages.bib]
biblio-style: apalike
link-citations: yes
editor_options: 
  chunk_output_type: console
---

@@

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
), "packages.bib")

# Set rounding
op <- options()
options(digits = 2)
```

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 sec) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004].

Describing individual differences in short sprint performance, quantifying training intervention effects, but more importantly, understanding short sprint *bottle necks* (i.e. limiting factors) has been one of the prominent tasks of sports science profession since its establishment. The \pkg{shorts} package [@R-shorts], written in the R language [@R-base], represents an open-source tool that aims to help sports scientist in achieving these goals through mathematical modeling.

In the current paper, we will provide explanation of one commonly used mathematical equation to model short sprints, modeling applications using the \pkg{shorts} package, issues that can arise during measurement and estimation, and potential solutions to those problems.

# Mathematical model

Short sprints have been modeled using the mono-exponential equation \@ref(eq:velocity-time) originally proposed by @doi:10.1098/rspb.1927.0035, and more recently popularized by @clarkNFLCombine40Yard2017, and @samozinoSimpleMethodMeasuring2016. Equation \@ref(eq:velocity-time) represents function for instantaneous horizontal velocity $v$ given the time $t$ and two model parameters:

@@

The parameters of the equation \@ref(eq:velocity-time) are \textit{maximum sprinting speed} (MSS; expressed in $ms^{-1}$) and \textit{relative acceleration} (TAU). Mathematically, TAU represents the ratio of MSS to initial acceleration (MAC; \textit{maximal acceleration}, expressed in $ms^{-2}$) \@ref(eq:maximal-acceleration).

@@

By derivating equation \@ref(eq:velocity-time), we can get equation for horizontal acceleration \@ref(eq:acceleration-time).

@@

By integrating equation \@ref(eq:velocity-time), we can get equation for distance covered \@ref(eq:distance-time).

@@

kinematics_long <- kinematics %>%
  pivot_longer(
    cols = c("distance", "velocity", "acceleration"),
    names_to = "variable"
  ) %>%
  mutate(
    variable = factor(
      variable,
      levels = c("distance", "velocity", "acceleration"),
      labels = c(
        expression("Distance (m"*")"),
        expression("Velocity (ms"^-1*")"),
        expression("Acceleration (ms"^-2*")"))
    )
  )

gg <- ggplot(
  kinematics_long,
  aes(x = time, y = value, color = Athlete)
) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(
    ~variable,
    scales = "free_y",
    labeller = label_parsed) +
  ylab(NULL) +
  xlab("Time (s)") +
  theme(legend.position = "top", legend.title = element_blank())

@@

If we plot acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get Acceleration-Velocity profile, which is linear, according to the mathematical model.

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(10) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2*")")) +
  xlab(expression("Velocity (ms"^-1*")")) +
  theme(legend.position = "top", legend.title = element_blank())

@@

# Estimation using shorts package

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40m), and (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in \pkg{shorts} package using non-linear least squares regression implemented in the `nls` function in the \pkg{base} R [@R-base] and `nlme` function in the \pkg{nlme} package [@R-nlme] for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 $ms^-1$, TAU equal to 1.3, and MAC equal to 6.92 $ms^-2$ performing 40m sprint with timing gates positioned at each 10m split. When it comes to split times, distance is a predictor, and time is outcome variable, thus the equation \@ref(eq:velocity-time) takes another form:

@@

$W$ in equation \@ref(eq:velocity-time) represents Lambert's W function [@R-LambertW]. MSS and TAU parameters are estimated using `shorts::model_using_splits` function:

@@

@@

Maximal relative power (PMAX) from the output is estimated using $\frac{MSS \times MAC}{4}$, which disregards the air resistance. `time_correction` and `distance_corection` parameters will be covered later in the article.

Besides providing *residual standard error* (RSE), \pkg{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls` function:

@@

Here is a plot of observed and model predicted split times, with the help of the \pkg{ggplot2} package [@R-ggplot2]:

@@

ggplot(df, aes(x = distance)) +
  theme_bw(10) +
  geom_line(aes(y = time)) +
  geom_point(aes(y = pred_time), color = "red") +
  xlab("Distance (m)") +
  ylab("Time (s)")
```

@@

To estimate force production at distance or time (using `shorts::predict_force_at_distance` and `shorts::predict_force_at_time` functions), as well as power production (using `shorts::predict_power_at_distance` and `shorts::predict_power_at_time` functions), one needs to take into account the air resistance. Air resistance (N) is estimated using `shorts::get_air_resistance` function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature ($C^\circ$), and wind velocity ($ms^-1$) as parameters (please refer to @arsacModelingEnergetics100m2002, @samozinoSimpleMethodMeasuring2016, and @vaningenschenauCanCyclePower1991 for more information):

@@

When estimating force and power, the air resistance parameters can be set using `"..."`, which are forwarded to the `shorts::get_air_resistance`:

@@

The easiest way to get all kinematics for short sprints is to use `shorts::predict_kinematics` functions:

@@

Plotting the model predictions can be done once we convert data from wide to long with the help of \pkg{dplyr} [@R-dplyr], \pkg{tidyr} [@R-tidyr], and \pkg{tidyverse} [@R-tidyverse] packages:

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90% of maximum sprinting speed is reached, or where peak power is within 90% range. To identify these values, \pkg{shorts} package comes with `shorts::find_` family of functions:

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use `shorts::mixed_model_using_splits` function. To demonstrate this functionality, we load the `split_times` dataset provided in the \pkg{shorts} package:

@@

The following plot contains kinematics for all athletes in `split_times` dataset. Please note that power calculation takes default parameters for each individual:

@@

ggplot(
  filter(df, distance < 40),
  aes(x = distance, y = value, group = athlete, color = athlete)
) +
  theme_bw(10) +
  facet_wrap(~name, scales = "free_y") +
  geom_line(alpha = 0.7) +
  ylab(NULL) +
  xlab("Distance (m)") +
  theme(legend.position = "top", legend.title = element_blank())
```

@@

Estimation of the short sprint profile using radar gun data takes time as predictor and velocity as an target or outcome variable. Thus equation \@ref(eq:velocity-time) is used to estimate MSS and TAU.

Let's consider the same example of an athlete with MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$ms^-2$ performing 40m sprint with velocity estimated using radar run (in this case with 1Hz sampling rate).

@@

Mixed-effects model using radar data is done using `shorts::mixed_model_using_radar` function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

@@

There is a challenge when collecting sprint data that could have a substantial impact on modeled outcomes. To ensure accurate parameter outcomes, the initial force production must be synced with start time. Below we describe this challenge when using radar guns or timing gates and suggest potential solutions within the \pkg{shorts} package.

@@

One source of error in the modeled estimation using radar gun is the time synchronization. In theory, synchronization is ideal when sprint is initiated at $t=0$ (i.e., $v(t=0) = 0$). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5second to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

ggplot(plot_df, aes(x = value, y = velocity, color = `Sync issue`)) +
  theme_bw(10) +
  geom_line(alpha = 0.7) +
  xlab("Time (s)") +
  ylab(expression("Velocity ("*ms^-1*")"))
```

@@

As can be seen from the example, TAU is affected by an error in synchronization of time with velocity. The potential solution incorporated into the \pkg{shorts} package involves estimation of the *time correction* parameter using the following equation:

@@

This model is incorporated in the `shorts::model_using_radar_with_time_correction` function:

@@

When it comes to mixed-model approach, time correction can be modeled as a fixed effect or random effect using the `shorts::mixed_model_using_radar_with_time_correction` function.

@@

Let's imagine we have two twin brothers with same short sprint characteristics: MSS equal to 9$ms^-1$, TAU equal to 1.3, and MAC equal to 6.92$ms^-2$. Let's call them John and Jack. The are both performing 40m sprint using timing gates set at 5, 10, 20, 30, and 40m. The initial timing gate at the start (i.e., $d=0m$) serves to activate the timing system (i.e., when they cross the beam).

John represents the *theoretical model*, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a *practical model*, and decides to move slightly behind the initial timing gate (i.e. for 0.5m) and use body rocking to initiate the sprint start. In other words, Jack is using a *flying start*, a common scenario when testing field sports athletes.  Let's see how their sprints differ.

@@

And here is a graphical representation of the sprint splits:

```{r}
plot_df <- split_times %>%
  select(distance, john_time, jack_time) %>%
  rename(John = john_time, Jack = jack_time) %>%
  pivot_longer(cols = -1, names_to = "athlete", values_to = "time") %>%
  mutate(distance = factor(distance))

ggplot(plot_df, aes(x = distance, y = time, color = athlete, group = athlete)) +
  theme_bw() +
  geom_point() +
  geom_line() +
  xlab("Distance (m)") +
  ylab("Time (s)")
```

Using the following code, we can see how different are the estimated MSS and TAU parameters:

@@

As can be seen from the results, flying starts yields biased estimates, particularly for the TAU, MAC and PMAX.

Below is a simulation sprint with 5, 10, 20, 30, 40, and 50m splits, with MSS and MAC varying from 6 to 9 ($ms^-1$ and $ms^-2$ respectively), and flying start distance varying from 0 to 1m.

```{r}
sim_df <- expand.grid(
  MSS = c(6, 7, 8, 9),
  MAC = c(6, 7, 8, 9),
  flying_start_distance = c(
    seq(0, 0.001, length.out = 20),
    seq(0.001, 0.01, length.out = 20),
    seq(0.01, 0.1, length.out = 20),
    seq(0.1, 1, length.out = 20)
  ),
  distance = c(5, 10, 20, 30, 40, 50)
)

sim_df <- sim_df %>%
  mutate(
    TAU = MSS / MAC,
    PMAX = MSS * MAC / 4,
    true_distance = distance + flying_start_distance,
    true_time = shorts::predict_time_at_distance(true_distance, MSS, TAU),
    stolen_time = shorts::predict_time_at_distance(flying_start_distance, MSS, TAU),
    time = true_time - stolen_time
  )

@@

Now when we have a simulation dataset, we can check the model estimates and predictions, given the flying start distance:

@@

# estimated parameters and predited time
model_df <- sim_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  do(pred_wrapper(.)) %>%
  ungroup()

@@

The following image demonstrates the effect of flying start distance on estimated MSS:

```{r}
# Estimates plot
df <- model_df %>%
  group_by(MSS, TAU, flying_start_distance) %>%
  slice(1) %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2))
  )

# MSS
ggplot(df, aes(x = flying_start_distance, y = est_MSS, color = MAC_string)) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~MSS_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MSS (m/s)") +
  theme(legend.title = element_blank())
```

As can be seen from the image, MSS is underestimated as flying start distance increases. The following image demonstrates the effect of flying start distance on estimated MAC:

```{r}
# MAC
ggplot(df, aes(x = flying_start_distance, y = est_MAC, color = MSS_string)) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated MAC (m/s/s)") +
  theme(legend.title = element_blank())
```

MAC (and also TAU) are highly affected by the flying start distance, and from the figure we can notice that MAC is overestimated as cheat distance increases.

And finally, the following image demonstrates the effect of flying start distance on estimated PMAX:

```{r}
# PMAX
ggplot(df, aes(x = flying_start_distance, y = est_PMAX, color = MSS_string)) +
  theme_bw() +
  geom_line(alpha = 0.7) +
  facet_wrap(~MAC_string, scales = "free_y") +
  xlab("Flying start distance (m)") +
  ylab("estimated PMAX (W/kg)") +
  theme(legend.title = element_blank())
```

@@

Model residuals are also affected by flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40m versus 5, 15, 30m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simulation:

```{r}
# Residuals
model_df <- model_df %>%
  mutate(
    MSS_string = paste("MSS =", MSS),
    TAU_string = paste("TAU =", TAU),
    MAC_string = paste("MAC = ", round(MAC, 2)),
    PMAX_string = paste("PMAX = ", round(PMAX, 2)),
    group = paste(MSS, MAC, flying_start_distance)
  )

ggplot(model_df, aes(y = residuals, x = distance, color = flying_start_distance, group = group)) +
  theme_bw() +
  geom_line(alpha = 0.3) +
  facet_grid(MSS_string ~ MAC_string) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)")
```

If we merge individual facets (i.e., combinations of MSS and MAC), we can get simpler images conveying issues with residuals when there is a flying start:

```{r}
ggplot(model_df, aes(y = residuals, x = distance, color = flying_start_distance, group = group)) +
  theme_bw() +
  geom_line(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_gradientn(colours = terrain.colors(5, rev = FALSE)) +
  xlab("Distance (m)") +
  ylab("Predicted time - observed time (s)")
```

Clearly, any type of flying start where there is a difference between initial force production and start time can result in biased parameters and predictions. Since maximal sprint speed is difficult to improve, the effects of start inconsistencies can mask effects of the training intervention. It is thus crucial to standardize the start when testing and implementing the following techniques when using the \pkg{shorts} package.

### How to overcome missing the initial force production when using timing gates?

A potential solution is to use a correction factor - the recommendation in the literature is +0.5sec [@haugenSprintMechanicalProperties2020; @haugenSprintMechanicalVariables2019]. Interestingly, the average difference between using timing gates and a block start for 40 m sprint time was 0.27 sec [@haugenDifferenceStartImpact2012]. So, while a timing correction factor is warranted to avoid subsequent errors in estimates of kinetic variables (e.g., overestimate power), a correction factor that is too large will have the opposite effect (e.g., underestimate power).

When implementing *time correction*, equation \@ref(eq:time-distance) gets another form:

@@

Rather than providing *apriori* time correction from the literature, \pkg{shorts} package provides estimation of this parameter from the data provide, together with MSS and TAU. To estimate time correction, we use `shorts::model_using_splits_with_time_correction` function. Here is how we can estimate Jack parameters using either provided time correction (e.g., +0.3 and +0.5s) or estimated time correction:

```{r}
jack_profile_fixed_time_short <- shorts::model_using_splits(
  distance = split_times$distance,
  time = split_times$jack_time,
  time_correction = 0.3
)

jack_profile_fixed_time_long <- shorts::model_using_splits(
  distance = split_times$distance,
  time = split_times$jack_time,
  time_correction = 0.5
)

@@

jack_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters),
  unlist(jack_profile_fixed_time_short$parameters),
  unlist(jack_profile_fixed_time_long$parameters),
  unlist(jack_profile_time_estimated$parameters)
)

rownames(jack_parameters) <- c(
  "John",
  "Jack - No corrections",
  "Jack - Fixed time correction (+0.3s)",
  "Jack - Fixed time correction (+0.5s)",
  "Jack - Estimated time correction"
)

@@

In Jack's case, both +0.3s fixed time correction and time correction estimation yield parameters closer to John's (i.e. true parameters).

Another model definition, which is novel approach implemented in the \pkg{shorts} packages, is to ulitize *distance correction*, besides time correction. Thus, equation \@ref(eq:time-distance) gets another form:

@@

jack_parameters <- rbind(
  unlist(john_profile$parameters),
  unlist(jack_profile$parameters),
  unlist(jack_profile_fixed_time_short$parameters),
  unlist(jack_profile_fixed_time_long$parameters),
  unlist(jack_profile_time_estimated$parameters),
  unlist(jack_profile_distance_correction$parameters)
)
